---
output: 
  github_document:
    df_print: kable
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%",
  fig.align = "center",
  dpi = 300
)
options(scipen = 999, digits = 4)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(formatC(x, big.mark = ",", format = "fg", digits = 2))
    } else{
      return(x)
    }
  }
)
```

```{r}
# libraries
library(tidyverse)
library(here)
library(knitr)
```


Load hybrid article-level dataset and journal classification

```{r}
jn_subjects <- read_csv(here::here("data", "jn_subjects.csv"))
hybrid_df <- readr::read_csv(here::here("data", "hybrid_articles.csv"))
subject_hybrid <- left_join(hybrid_df, jn_subjects, by = "issn") %>%
  select(doi, issn, top_level, subject_area)
```


## Subject field (top-level)

Number of Journals (full and fractional counting)

```{r}
subject_hybrid %>% 
  group_by(issn, top_level) %>%
  summarise(n = n()) %>% 
  mutate(fractions = n / sum(n)) %>% 
  ungroup() %>% 
  group_by(top_level) %>% 
  summarise(journals_n = n_distinct(issn),
            journals_frac = sum(fractions)) %>% 
  arrange(desc(journals_frac))
```
  
Articles

```{r}
subject_hybrid %>% 
  group_by(doi, top_level) %>%
  summarise(n = n()) %>% 
  mutate(fractions = n / sum(n)) %>% 
  ungroup() %>% 
  group_by(top_level) %>% 
  summarise(articles_n = n_distinct(doi),
            articles_frac = sum(fractions)) %>% 
  arrange(desc(articles_frac))
```

### Subject area

Number of Journals (full and fractional counting)

```{r}
subject_hybrid %>% 
  group_by(issn, subject_area) %>%
  summarise(n = n()) %>% 
  mutate(fractions = n / sum(n)) %>% 
  ungroup() %>% 
  group_by(subject_area) %>% 
  summarise(journals_n = n_distinct(issn),
            journals_frac = sum(fractions)) %>% 
  arrange(desc(journals_frac))
```

Articles

```{r}
subject_hybrid %>% 
  group_by(doi, subject_area) %>%
  summarise(n = n()) %>% 
  mutate(fractions = n / sum(n)) %>% 
  ungroup() %>% 
  group_by(subject_area) %>% 
  summarise(articles_n = n_distinct(doi),
            articles_frac = sum(fractions)) %>% 
  arrange(desc(articles_frac))
```

## Agreement only

```{r}
jn_subjects <- read_csv(here::here("data", "jn_subjects.csv"))
hybrid_df <- readr::read_csv(here::here("data", "hybrid_articles.csv")) %>%
  filter(oa_sponsor_type == "FundingBody")
subject_agreement_hybrid <- left_join(hybrid_df, jn_subjects, by = "issn") %>%
  select(doi, issn, top_level, subject_area)
```


## Subject field (top-level)

Number of Journals (full and fractional counting)

```{r}
subject_agreement_hybrid %>% 
  group_by(issn, top_level) %>%
  summarise(n = n()) %>% 
  mutate(fractions = n / sum(n)) %>% 
  ungroup() %>% 
  group_by(top_level) %>% 
  summarise(journals_n = n_distinct(issn),
            journals_frac = sum(fractions)) %>% 
  arrange(desc(journals_frac))
```
  
Articles

```{r}
subject_agreement_hybrid %>% 
  group_by(doi, top_level) %>%
  summarise(n = n()) %>% 
  mutate(fractions = n / sum(n)) %>% 
  ungroup() %>% 
  group_by(top_level) %>% 
  summarise(articles_n = n_distinct(doi),
            articles_frac = sum(fractions)) %>% 
  arrange(desc(articles_frac))
```

### Subject area

Number of Journals (full and fractional counting)

```{r}
subject_agreement_hybrid %>% 
  group_by(issn, subject_area) %>%
  summarise(n = n()) %>% 
  mutate(fractions = n / sum(n)) %>% 
  ungroup() %>% 
  group_by(subject_area) %>% 
  summarise(journals_n = n_distinct(issn),
            journals_frac = sum(fractions)) %>% 
  arrange(desc(journals_frac))
```

Articles

```{r}
subject_agreement_hybrid %>% 
  group_by(doi, subject_area) %>%
  summarise(n = n()) %>% 
  mutate(fractions = n / sum(n)) %>% 
  ungroup() %>% 
  group_by(subject_area) %>% 
  summarise(articles_n = n_distinct(doi),
            articles_frac = sum(fractions)) %>% 
  arrange(desc(articles_frac))
```
